## colonization limitation

Some consideration of colonization limitation, specifically between-patch transmission of infection. Such transmission is (we imagine) *jointly* limited by the number of infectious fleas, and the number of places available in emigrating rats.

We will say the number of fleas ($F$) is proportional to the total number of disease-induced host deaths in a given epidemic period, while the number of places in emigrating rats ($R$) is proportional to the number of emigrating rats. Yuyang originally wrote the realized transmission as $\textrm{min}(R,F)$.

How do we think about this? Per capita success of emigrating fleas is a saturating exponential of the rat/flea ratio ...

We should instead consider $t(F,R) = F \left(1 - \exp(-R/F)\right)$. This has the desired properties:

* $t(0,R) = t(F,0) = 0$
* $\lim_{F\to\infty} t(F,R) = R$
* $\lim_{R\to\infty} t(F,R) = F$
* derivatives with respect to $F$ and $R$ are always positive (as long as $F, R > 0$)

It's a little bit hard to get my head around the fact that this is *asymmetric*

```{r curve}
library(emdbook)
cc <- curve3d(F*(1-exp(-R/F)), varnames = c("F", "R"),
        to = c(50, 50), sys3d = "image", n = 60)
```
It's certainly not *very* asymmetric ...


```{r}
matplot(cbind(cc$z[,2], cc$z[2,]), type = "l")
```


```{r}
matplot(cbind(cc$z[,50], cc$z[50,]), type = "l")
```

Compare with min:

```{r}
cc2 <- curve3d(min(F,R), varnames = c("F", "R"),
        to = c(50, 50), sys3d = "image", n = 60)
```
