---
title: "Why does post-establishment fade-out increase with $R_0$?"
date: 2026-10-06
format:
  html:
    embed-resources: true
execute:
  echo: false
  message: false
  warning: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(knitr)
library(here)

batch_dir <- here::here("fadeout", "output", "occupancy_factorial_100y_batch",
                        "data")
combo <- read.csv(file.path(batch_dir, "combo_summary.csv"))
runs <- read.csv(file.path(batch_dir, "run_summary.csv"))

source(here::here("fadeout", "logistic_burnout",
                  "logistic_burnout_functions.R"))

gamma <- 0.2          ## per day (one generation = 5 days)
r <- 0.125 * gamma    ## logistic growth rate per day
threshold <- 1.5      ## establishment threshold, in units of T0
```

## The observation

In the batch runs (`occupancy_factorial_100y_batch.R`), a patch counts as
*established* once its episode has lasted $1.5\,T_0$, where $T_0$ is the
intrinsic period of the damped endemic oscillation. Under constant
transmission, the hazard of fade-out after establishment rises steeply with
$R_0$ at every $K$:

```{r hazard-table}
combo |>
  filter(scenario == "constant", alpha == 1e-4, K %in% c(1000, 3000, 5000, 10000)) |>
  select(R0, K, fadeout_hazard) |>
  pivot_wider(names_from = K, values_from = fadeout_hazard,
              names_prefix = "K = ") |>
  kable(digits = 5,
        caption = "Fade-outs per established patch-year (constant transmission, alpha = 1e-4/day, 5 replicates pooled)")
```

## Candidate explanations

1. **Transient ringing after establishment.** An established patch is still
   recovering from the initial overshoot. At higher $R_0$ the overshoot is
   larger and is damped more slowly, so the 2nd and 3rd troughs are deeper.
   This is the $P_2, P_3, \ldots$ effect: the probability of surviving the
   *second and later* troughs, not just the first. It predicts fade-outs
   concentrated at young ages (a few $T_0$).
2. **Stochastic amplification near equilibrium.** Long after the transient has
   gone, demographic noise keeps exciting the weakly damped oscillatory mode
   (quasi-cycles; Alonso, McKane & Pascual 2007). Weaker damping means larger
   sustained fluctuations and occasional troughs deep enough to lose the
   infection. This predicts a roughly constant hazard at all ages.
3. **Smaller endemic prevalence.** A lower $I^*$ at higher $R_0$ would raise
   the hazard. This is ruled out: $I^*/K = (r/\gamma)(1 - 1/R_0)/R_0$ is
   nearly flat across the range of $R_0$ (table below).

A minor additional factor is that $T_0$ is shorter at high $R_0$, so there are
more troughs per year. That accounts for at most a factor of about 1.5, not
orders of magnitude.

## Linear analysis

Linearizing the logistic SI model about its endemic equilibrium gives a
damped oscillation with damping rate $r/(2R_0)$ and natural frequency
$\omega_0^2 = \gamma r (R_0 - 1)/R_0 - r^2/(4R_0^2)$. Damping *per cycle*
weakens as $R_0$ increases, because the decay rate falls while the frequency
rises:

```{r linear-table}
linear <- tibble(R0 = c(1.5, 2, 2.5, 3, 3.5)) |>
  mutate(
    damping = r / (2 * R0),
    omega0 = sqrt(gamma * r * (R0 - 1) / R0 - r^2 / (4 * R0^2)),
    T0 = 2 * pi / omega0,
    retained = exp(-damping * T0),
    I_star_K = (r / gamma) * (1 - 1 / R0) / R0
  )
linear |>
  kable(digits = c(1, 4, 4, 0, 2, 4),
        col.names = c("R0", "damping rate (/day)", "omega0 (/day)",
                      "T0 (days)", "amplitude retained per cycle", "I*/K"))
```

Both explanation 1 and explanation 2 follow from this weakening damping. A
slowly damped transient rings for longer, and the variance of noise-driven
quasi-cycles scales inversely with the damping rate.

## Young versus older fade-outs

To separate the two mechanisms, split the time after establishment into a
young window ($1.5$–$3\,T_0$, around the second trough) and everything later.
Then compute the hazard in each window (constant transmission; $K$ = 5000 and
10000; all $\alpha$ pooled). Time at risk in the young window comes from the
episode-duration histograms ($0.1\,T_0$ bins, using bin midpoints).

```{r young-old}
young_end <- 3

hist <- readRDS(file.path(batch_dir, "duration_hist.rds"))

sel <- runs |>
  filter(scenario == "constant", K %in% c(5000, 10000)) |>
  select(run_id, R0, T0, n_fadeout, established_years)

young <- hist |>
  inner_join(sel, by = "run_id") |>
  mutate(age = (bin + 0.5) * 0.1) |>   ## bin midpoint, in units of T0
  filter(age > threshold) |>
  group_by(run_id) |>
  summarise(
    young_fadeouts = sum(n[class == "fadeout" & age < young_end]),
    young_years = sum(n * (pmin(age, young_end) - threshold) * T0) / 365
  )

age_hazard <- sel |>
  left_join(young, by = "run_id") |>
  mutate(across(c(young_fadeouts, young_years), \(x) coalesce(x, 0))) |>
  group_by(R0) |>
  summarise(
    young_hazard = sum(young_fadeouts) / sum(young_years),
    old_hazard = sum(n_fadeout - young_fadeouts) /
      sum(established_years - young_years),
    frac_old = sum(n_fadeout - young_fadeouts) / sum(n_fadeout),
    n_fadeout = sum(n_fadeout)
  ) |>
  mutate(ratio = if_else(young_hazard > 0, young_hazard / old_hazard, NA))

age_hazard |>
  select(R0, young_hazard, old_hazard, ratio, frac_old, n_fadeout) |>
  kable(digits = c(1, 5, 6, 0, 2, 0),
        col.names = c("R0", "hazard, 1.5-3 T0 (/yr)", "hazard, > 3 T0 (/yr)",
                      "ratio", "fraction of fade-outs > 3 T0",
                      "total fade-outs"))
```

* Per unit time, the risk in the young window is 20–80 times the later risk,
  so transient ringing (explanation 1) is real.
* The later hazard also rises by more than two orders of magnitude between
  $R_0 = 2$ and $3.5$. Patches spend much more time at older ages, so most
  fade-outs happen there. Stochastic amplification (explanation 2) therefore
  contributes most of the total.
* The $R_0 = 1.5$ row rests on very few fade-outs and is mostly noise.

## Deterministic-trough calculation of $P_2$, $P_3$

The `burnout` package (Earn & Bolker) computes $\mathcal{P}_m$ (`Pm_prob()`)
only for the SIR model with births and deaths. Our within-patch model is an
SI model with disease-induced death and logistic host growth, so we use the
same method as implemented for that model in
`fadeout/logistic_burnout/logistic_burnout_functions.R`
(`logistic_multitrough_probabilities()`). The method has three steps:

1. Follow the deterministic trajectory from a single infective in a patch at
   $S = K$.
2. At the entry to each trough's boundary layer, compute the Kendall
   extinction probability $q$ of one lineage while susceptibles recover
   logistically.
3. Raise $q$ to the number of infectives at entry.

$Q_j$ is the probability of burnout in trough $j$, given survival through
trough $j-1$. The table also shows when each trough is entered, in units of
$T_0$ since introduction.

```{r multitrough}
mt_grid <- expand_grid(R0 = c(2, 2.5, 3, 3.5), K = c(3000, 5000, 10000))

T0_days <- setNames(linear$T0, linear$R0)

multitrough <- mt_grid |>
  rowwise() |>
  reframe({
    tt <- logistic_multitrough_probabilities(R0, 0.125, K,
                                             max_troughs = 3)$trough_table
    tibble(R0 = R0, K = K, trough = tt$trough_index,
           ## t_in is in generations; one generation is 1/gamma days
           t_in_T0 = tt$t_in / gamma / T0_days[as.character(R0)],
           Q = tt$Q_burnout)
  })
```

To compare with the simulations, take colonization episodes that are still
infected at $1.5\,T_0$ (constant transmission, all $\alpha$ pooled). The
comparable simulated quantity is the proportion of these episodes that go
extinct before $3\,T_0$, a window that contains the second trough.

```{r multitrough-compare}
empirical <- hist |>
  filter(!initial) |>
  inner_join(runs |> filter(scenario == "constant") |> select(run_id, R0, K),
             by = "run_id") |>
  mutate(age = (bin + 0.5) * 0.1) |>
  filter(age > threshold) |>
  group_by(R0, K) |>
  summarise(
    sim_p = sum(n[class == "fadeout" & age < young_end]) / sum(n),
    at_risk = sum(n),
    .groups = "drop"
  )

multitrough |>
  pivot_wider(id_cols = c(R0, K), names_from = trough,
              values_from = c(t_in_T0, Q), names_sep = "") |>
  select(R0, K, t_in_T02, t_in_T03, Q1, Q2, Q3) |>
  left_join(empirical, by = c("R0", "K")) |>
  arrange(K, R0) |>
  mutate(across(Q1:Q3, \(x) formatC(x, format = "g", digits = 2))) |>
  kable(digits = 3,
        col.names = c("R0", "K", "trough 2 entry (T0)", "trough 3 entry (T0)",
                      "Q1", "Q2", "Q3",
                      "simulated P(extinct 1.5-3 T0)", "episodes at risk"))
```

* The deterministic-trough theory reproduces the qualitative pattern: $Q_2$
  rises by many orders of magnitude with $R_0$. At $K = 5000$, for example,
  it goes from about $10^{-6}$ at $R_0 = 2$ to 0.5 at $R_0 = 3.5$.
  Trough 2 starts at about $1.7$–$2.2\,T_0$, just after the establishment
  threshold, so explanation 1 explains the steep young-age hazard.
* Quantitatively the fit is poor in both directions:
  * At low $R_0$ the theory is far too small. It has no noise, so it misses
    fade-outs from explanation 2 and from lineages lingering after the first
    trough.
  * At high $R_0$ it is too large, probably because colonizations in the
    metapopulation often reach patches whose hosts have not yet recovered to
    $K$ after an earlier epidemic. The overshoot is then smaller than from
    $S = K$.
* The "Q" values describe a single deterministic pass and ignore the
  subsequent stationary hazard entirely. They therefore cannot account for the
  70–80% of fade-outs that happen after $3\,T_0$.

## Summary

The fade-out hazard after establishment rises with $R_0$ because damping per
cycle of the within-patch oscillation weakens as $R_0$ rises: the amplitude
retained per cycle goes from 0.27 at $R_0 = 1.5$ to 0.69 at $R_0 = 3.5$.
Weaker damping does two things:

* the post-invasion transient keeps ringing, giving deep second and third
  troughs (high $Q_2$, $Q_3$);
* demographic noise sustains larger quasi-cycles around equilibrium.

The first effect dominates the hazard per unit time early on. The second
dominates the total number of fade-outs.
