---
title: more notes on metapop patch occupancy dynamics
---

How did we calibrate final size earlier, when we were doing it by brute force fo r two-strain models? See `two_strain/ode_test_proc.R`: quadratic polynomial. Here, I think we might be better off with a spline fit.

```{r}
library("here")
library("dplyr")
library("ggplot2"); theme_set(theme_bw())
library("patchwork")
library("mgcv")
dd <- (readRDS(here::here("odin/sharcnet/outputs/",
                          "euler_onepatch_onestrain_extinct_logistic_continuous.rds"))
    |> select(R0, K, ext_prob.I1)
)
```

```{r plot}
gg0 <- ggplot(dd, aes(R0, K, fill = ext_prob.I1)) +
    geom_raster() +
    scale_y_log10(expand = c(0,0)) +
    scale_fill_viridis_c() +
    scale_x_continuous(expand = c(0,0))
```

```{r}
## using default 'cr' basis; increased k as much as I can given grid resolution
gam_fit <- gam(ext_prob.I1 ~ te(R0, K, k = c(12, 12)),data = dd)
dd <- within(dd, {
    fit <- predict(gam_fit)
    err <- fit - ext_prob.I1
})
gg_fit <- gg0 + dd + aes(fill = fit)
gg_err <- gg0 + dd + aes(fill = err)
gg0 + gg_fit + gg_err
```

Good enough?

Could store this as an object and use it to predict patch occupancy dynamics

Once we have this, I think we should be able to replicate the patch occupancy curve pretty well from first principles.

Much harder (and more of intellectual than practical interest); repeat perturbation analysis from burnout paper for logistic host dynamics ...

